\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs}

\title{Depth Shortcut Predictor for DiT: Training Summary}
\date{}

\begin{document}
\maketitle

\section{Setup}
Let a DiT backbone have depth $L=12$, $P=256$ patch tokens, hidden size
$D=768$, and $T=50$ discrete training timesteps. For a noised latent
$x_t$ at timestep index $q\in\{0,\ldots,49\}$, a full DiT forward returns
\[
    \hat v,\quad Z_0,Z_1,\ldots,Z_L,\quad e_t^{\mathrm{DiT}},
\]
where $Z_0$ is the post-patch-embedding state and $Z_l\in\mathbb{R}^{B\times P\times D}$
is the state after block $l$.

All shortcut supervision is done in tokenwise direction space:
\[
    U_l=\operatorname{dir}(Z_l)
    =\frac{Z_l}{\|Z_l\|_2+\epsilon},
    \qquad U_l\in\mathbb{R}^{B\times P\times D}.
\]
The shortcut predictor receives a source direction, source layer, target
layer, and DiT timestep embedding:
\[
    Y_{a\rightarrow b}=P_\theta(U_a,a,b,t_q),\qquad
    \hat U_{a\rightarrow b}=\operatorname{dir}(Y_{a\rightarrow b}).
\]

\section{Conditioning}
For $0\leq a<b\leq L$, let $\Delta=b-a$. The predictor uses learned layer
embeddings of dimension $16$:
\[
    h_a=W_{\mathrm{state}}o_a,\quad
    h_b=W_{\mathrm{state}}o_b,\quad
    h_\Delta=W_\Delta o_\Delta .
\]
The depth condition is
\[
    h_{\mathrm{depth}}=[h_a,h_b,h_b-h_a,h_\Delta],
\]
and the predictor condition is produced by an MLP from
\[
    c_{\mathrm{in}}=
    [h_{\mathrm{depth}},\operatorname{sg}(e_t^{\mathrm{DiT}})].
\]
The stop-gradient prevents shortcut training from directly perturbing the
backbone timestep embedding through the condition path.

\section{Predictor Architecture}
The source direction is still the main input, but the predictor also receives
a stop-gradient log-norm context map. For source state $Z_a$, define
\[
    r_a(i)=\|Z_a(i)\|_2,\qquad
    U_a(i)=\frac{Z_a(i)}{r_a(i)+10^{-6}},\qquad
    m_a(i)=\ln(r_a(i)+10^{-6}).
\]
For SiT-B/2, two norm-map channels are formed:
\[
    M_a^{\mathrm{abs}}(i)=\frac{m_a(i)-5.5}{1.5},
\]
\[
    M_a^{\mathrm{rel}}(i)=
    \frac{m_a(i)-\operatorname{mean}_j m_a(j)}
    {\operatorname{std}_j m_a(j)+10^{-6}}.
\]
The map
\[
    M_a=[M_a^{\mathrm{abs}},M_a^{\mathrm{rel}}]
    \in\mathbb{R}^{B\times16\times16\times2}
\]
is stop-gradient before entering the shortcut predictor. A small norm encoder
produces
\[
    E_m=\mathrm{NormEncoder}(\operatorname{sg}(M_a))
    \in\mathbb{R}^{B\times16\times16\times32},
\]
with
\[
    \mathrm{NormEncoder}:
    \mathrm{Conv}_{1\times1}(2\to32)\rightarrow
    \mathrm{GELU}\rightarrow
    \mathrm{DWConv}_{3\times3}(32)\rightarrow
    \mathrm{GELU}.
\]
The source direction is reshaped from $B\times256\times768$ to
$B\times16\times16\times768$ and concatenated with the norm encoding:
\[
    X_a=[U_a^{\mathrm{grid}},E_m]\in\mathbb{R}^{B\times16\times16\times800}.
\]
The predictor input projection is therefore
\[
    X_a\rightarrow \mathrm{Linear}(800\to C),
\]
where $C$ is the predictor width. A ConvNeXt-style residual block is
\[
    h\leftarrow h+
    \mathrm{PW}_2\!\left(
    \mathrm{GELU}\!\left(
    \mathrm{PW}_1\!\left(
    \mathrm{DWConv}_{3\times3}(\mathrm{AdaLN}(h,c))
    \right)\right)\right),
\]
where
\[
    \mathrm{AdaLN}(h,c)=(1+\gamma(c))\mathrm{LN}(h)+\beta(c).
\]
The output projection predicts a residual direction update with a learnable
scalar output gate initialized at $\gamma_{\mathrm{out}}=0.05$:
\[
    Y_{a\rightarrow b}=U_a+\gamma_{\mathrm{out}}\Delta Y.
\]

\paragraph{Direction-only compatibility.}
In direction-only training and skip-FID, the predictor still receives
\[
    m_a=\ln(\|Z_a\|_2+10^{-6})
\]
so the shared backbone uses the same NormEncoder scale context. The caller can
set \texttt{return\_magnitude=False}; in that case the magnitude head is not
evaluated and only the direction output \(Y_{a\rightarrow b}\) is returned.

\begin{center}
\begin{tabular}{lrrrrl}
\toprule
Version & Width & Blocks & Expansion & Params & Use case \\
\midrule
P-Tiny  & 256 & 2 & 2 & 1.3--1.8M & fast debug \\
P-Small & 384 & 4 & 2 & 3.5--4.2M & lightweight runs \\
P-Base  & 768 & 4 & 2 & 10--11M & main balanced model \\
P-Large & 768 & 6 & 2 & 15--17M & long skips / ceiling runs \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Architecture Ablations}
Besides the original ConvNeXt predictor family, two parameter-matched
backbone variants are supported through the \texttt{--shortcut-predictor}
or \texttt{--predictor-variant} flag.

\paragraph{DilatedConvNeXt.}
This variant keeps the same AdaLN--ConvNeXt block and changes only the
depthwise convolution receptive field:
\[
    \mathrm{DWConv}_{3\times3}
    \quad\rightarrow\quad
    \mathrm{DWConv}_{3\times3,\ d=r}.
\]
The parameter count is essentially unchanged, so this isolates whether
long-gap shortcut quality is limited by local receptive field.

\begin{center}
\begin{tabular}{lrrrrl}
\toprule
Variant & Width & Blocks & Expansion & Dilation schedule & Params \\
\midrule
\texttt{dilated\_tiny}  & 256 & 2 & 2 & $[1,2]$ & 1.3--1.8M \\
\texttt{dilated\_small} & 384 & 4 & 2 & $[1,2,3,1]$ & 3.5--4.2M \\
\texttt{dilated\_base}  & 768 & 4 & 2 & $[1,2,3,1]$ & 10--11M \\
\texttt{dilated\_large} & 768 & 6 & 2 & $[1,2,3,1,2,3]$ & 15--17M \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Attention-Hybrid.}
This variant keeps local depthwise convolution and adds a lightweight
multi-head self-attention bottleneck over the $16\times16=256$ tokens:
\[
    \mathrm{AdaLN}\rightarrow
    \mathrm{DWConv}_{3\times3}\rightarrow
    \mathrm{MHSA}_{\mathrm{bottleneck}}\rightarrow
    \mathrm{PW\ MLP}\rightarrow\mathrm{residual}.
\]
To stay close to ConvNeXt parameter counts, the MLP expansion is reduced to
$1$ and the attention bottleneck uses approximately $C/2$ channels.
\[
    h\leftarrow h+
    \mathrm{PW}_2\!\left(
    \mathrm{GELU}\!\left(
    \mathrm{PW}_1\!\left(u+\gamma_{\mathrm{attn}}\mathrm{MHSA}(u)\right)
    \right)\right),
    \qquad \gamma_{\mathrm{attn}}=0.05 .
\]

\begin{center}
\begin{tabular}{lrrrrr}
\toprule
Variant & Width & Blocks & MLP exp. & Attn dim & Heads \\
\midrule
\texttt{attn\_hybrid\_tiny}  & 256 & 2 & 1 & 128 & 4 \\
\texttt{attn\_hybrid\_small} & 384 & 4 & 1 & 192 & 6 \\
\texttt{attn\_hybrid\_base}  & 768 & 4 & 1 & 384 & 8 \\
\texttt{attn\_hybrid\_large} & 768 & 6 & 1 & 384 & 8 \\
\bottomrule
\end{tabular}
\end{center}

The legacy names \texttt{tiny}, \texttt{small}, \texttt{base}, and
\texttt{large} map to \texttt{convnext\_tiny}, \texttt{convnext\_small},
\texttt{convnext\_base}, and \texttt{convnext\_large}. The training script
logs predictor, backbone, and total parameter counts before training starts.

\section{Direct Shortcut Loss}
Sample a direct pair by first sampling a gap $d\sim\mathrm{Uniform}\{1,\ldots,L\}$
and then $a\sim\mathrm{Uniform}\{0,\ldots,L-d\}$, with $b=a+d$.
The direct loss is cosine distance to the stop-gradient target direction:
\[
    \mathcal{L}_{\mathrm{dir}}
    =
    1-
    \frac{1}{BP}\sum_{n=1}^{B}\sum_{i=1}^{P}
    \hat U_{a\rightarrow b}^{(n)}(i)^\top
    \operatorname{sg}(U_b^{(n)}(i)).
\]

\section{Bootstrap Depth Consistency}
Sample a triplet $0\leq a<b<c\leq L$ by sampling positive gaps
$d_1=b-a$ and $d_2=c-b$ uniformly and resampling if $d_1+d_2>L$.
Two short shortcuts are composed:
\[
    \hat U_{a\rightarrow b}
    =
    \operatorname{dir}(P_\theta(U_a,a,b,t_q)),
\]
\[
    \hat U_{a\rightarrow b\rightarrow c}
    =
    \operatorname{dir}(P_\theta(\hat U_{a\rightarrow b},b,c,t_q)).
\]
The target is the EMA teacher's long shortcut:
\[
    \hat U_{a\rightarrow c}^{\mathrm{ema}}
    =
    \operatorname{sg}\!\left[
    \operatorname{dir}(P_{\bar\theta}(U_a,a,c,t_q))
    \right].
\]
The bootstrap loss is
\[
    \mathcal{L}_{\mathrm{boot}}
    =
    1-
    \frac{1}{BP}\sum_{n,i}
    \hat U_{a\rightarrow b\rightarrow c}^{(n)}(i)^\top
    \hat U_{a\rightarrow c}^{\mathrm{ema},(n)}(i).
\]
By default, bootstrap gradients can flow through the source state $U_a$ into
the DiT backbone. The CLI option
\texttt{--shortcut-bootstrap-detach-source} changes only this path:
\[
    U_a^{\mathrm{boot}}=\operatorname{sg}(U_a),\qquad
    m_a^{\mathrm{boot}}=\operatorname{sg}(m_a).
\]
Then $\mathcal{L}_{\mathrm{boot}}$ and $\mathcal{L}_{\mathrm{boot-mag}}$
update the predictor but do not update the DiT backbone through the bootstrap
source state. Direct shortcut loss and skip-FM behavior are unchanged.
The predictor teacher is updated after each optimizer step:
\[
    \bar\theta\leftarrow0.999\,\bar\theta+0.001\,\theta.
\]

\section{Direction--Magnitude Training Mode}
The optional \texttt{direction-magnitude} mode adds a log-norm transport head
and uses natural logarithms:
\[
    r_l(i)=\|Z_l(i)\|_2,\qquad
    m_l(i)=\ln(r_l(i)+10^{-6}).
\]
For the current SiT-B/2 setting, the measured token L2 range is roughly
$45.5$ to $1496.4$, corresponding to $\ln r\approx 3.82$ to $7.31$.
The head predicts a bounded normalized log-norm residual rather than a raw
norm. Use scale $s_m=3$:
\[
    \hat{\delta m}^{\mathrm{norm}}_{a\rightarrow b}(i)=\tanh(o_m(i)),
\]
\[
    \hat{\delta m}_{a\rightarrow b}(i)
    =
    s_m\,\hat{\delta m}^{\mathrm{norm}}_{a\rightarrow b}(i),
    \qquad
    \hat m_b(i)=m_a(i)+\hat{\delta m}_{a\rightarrow b}(i),
    \qquad
    \hat r_b(i)=\exp(\hat m_b(i)).
\]
The magnitude input uses two channels, not the raw log norm alone:
\[
    m_a^{\mathrm{abs}}(i)=\frac{m_a(i)-5.5}{1.5},
\]
\[
    m_a^{\mathrm{sp}}(i)=
    \frac{m_a(i)-\operatorname{mean}_{j}m_a(j)}
    {\operatorname{std}_{j}m_a(j)+10^{-6}}.
\]
Thus the magnitude head receives
\[
    M_a=[m_a^{\mathrm{abs}},m_a^{\mathrm{sp}}]\in\mathbb{R}^{B\times16\times16\times2}
\]
and applies
\[
    [h,M_a]\rightarrow \mathrm{AdaLN}(c)\rightarrow
    \mathrm{DWConv}_{3\times3}\rightarrow
    \mathrm{Linear}(C+2,C/4)\rightarrow\mathrm{GELU}\rightarrow
    \mathrm{Linear}(C/4,1).
\]
The final magnitude linear layer is zero-initialized. Its raw output $o_m$ is
bounded in normalized units:
\[
    \hat{\delta m}^{\mathrm{norm}}_{a\rightarrow b}=\tanh(o_m).
\]
The direct target is scaled and clipped:
\[
    \delta m_{a\rightarrow b}^{\mathrm{norm,tgt}}
    =
    \operatorname{clip}\left(\frac{m_b-m_a}{s_m},-1,1\right).
\]
The magnitude loss is bounded MSE on normalized log deltas:
\[
    \mathcal{L}_{\mathrm{mag}}
    =
    \frac{1}{2BP}\sum_{n,i}
    \left(
    \hat{\delta m}_{a\rightarrow b}^{\mathrm{norm},(n)}(i)
    -
    \operatorname{sg}\!\left(
    \delta m_{a\rightarrow b}^{\mathrm{norm,tgt},(n)}(i)
    \right)
    \right)^2.
\]
For bootstrap magnitude consistency, log residuals compose additively:
\[
    \delta m_{a\rightarrow c}
    =
    \delta m_{a\rightarrow b}+\delta m_{b\rightarrow c}.
\]
The loss compares the composed student residual to the EMA teacher residual:
\[
    \mathcal{L}_{\mathrm{boot-mag}}
    =
    \frac{1}{2BP}\sum_{n,i}
    \left(
    \hat{\delta m}_{a\rightarrow b}^{\mathrm{norm},(n)}(i)
    +\hat{\delta m}_{b\rightarrow c}^{\mathrm{norm},(n)}(i)
    -
    \operatorname{sg}\!\left(
    \hat{\delta m}_{a\rightarrow c}^{\mathrm{norm,ema},(n)}(i)
    \right)
    \right)^2.
\]

\paragraph{SiT-B/2 magnitude config.}
Use the following explicit config for the current SiT-B/2 experiments:
\[
\begin{array}{ll}
\text{log base} & \ln \\
\epsilon & 10^{-6} \\
s_m & 3.0 \\
m_a^{\mathrm{abs}} & (m_a-5.5)/1.5 \\
m_a^{\mathrm{sp}} & (m_a-\operatorname{mean}_j m_a(j))/(\operatorname{std}_j m_a(j)+10^{-6}) \\
\text{normalized target} & \operatorname{clip}((m_b-m_a)/3,-1,1) \\
\text{output bound} & \tanh(o_m) \\
\text{restore delta} & \hat{\delta m}=3\tanh(o_m) \\
\text{inference clamp} & \hat m_b\in[3,8] \\
\lambda_{\mathrm{mag}} & 0.375 \\
\lambda_{\mathrm{boot-mag}} & 0.1875
\end{array}
\]
Monitor $\operatorname{ClipRate}_{\delta m}=\mathbb{P}(|m_b-m_a|>3)$; if it
exceeds roughly $1\%$, rerun with $s_m=3.5$.

\section{Total Objective}
The backbone is still trained by the original flow matching loss:
\[
    \mathcal{L}_{\mathrm{FM}}
    =
    \mathbb{E}\|\hat v-(x_0-x_1)\|_2^2 .
\]
The joint objective is
\[
    \boxed{
    \mathcal{L}
    =
    \mathcal{L}_{\mathrm{FM}}
    +0.5\,\mathcal{L}_{\mathrm{dir}}
    +0.25\,\mathcal{L}_{\mathrm{boot}}
    }.
\]
In \texttt{direction-magnitude} mode, the objective becomes
\[
    \boxed{
    \mathcal{L}
    =
    \mathcal{L}_{\mathrm{FM}}
    +0.5\,\mathcal{L}_{\mathrm{dir}}
    +0.25\,\mathcal{L}_{\mathrm{boot}}
    +0.375\,\mathcal{L}_{\mathrm{mag}}
    +0.1875\,\mathcal{L}_{\mathrm{boot-mag}}
    }.
\]
Targets $U_b$ and $P_{\bar\theta}$ are stop-gradient. Gradients from the
shortcut losses can flow through the source state $U_a$ into the DiT backbone.

\section{Stochastic Skip-in-the-Loop Mode}
The optional \texttt{direction-magnitude-skip} mode reduces train--test
mismatch by occasionally feeding a predicted shortcut state into the remaining
backbone during training. First run the normal full forward to obtain
$Z_0,\ldots,Z_L$ and $\mathcal{L}_{\mathrm{FM}}$. Then, with probability
$p_{\mathrm{skip}}$, sample a skip gap from a left-truncated discrete normal:
\[
    d_{\mathrm{skip}}\in\{1,\ldots,10\},\qquad
    p(d)\propto
    \exp\left[-\frac{1}{2}\left(\frac{d-3}{2}\right)^2\right].
\]
This keeps $d=1,2,3$ likely, gives median near $3$, and leaves a low-probability
right tail for longer skips up to $10$. Then $b=a+d_{\mathrm{skip}}$ and
$a$ is sampled uniformly from valid source layers. Build the source using
detached hidden statistics for stability:
\[
    U_a^{\mathrm{skip}}=\operatorname{sg}(\operatorname{dir}(Z_a)),\qquad
    m_a^{\mathrm{skip}}=\operatorname{sg}(\ln(\|Z_a\|_2+10^{-6})).
\]
The predictor restores the skipped hidden state using its magnitude head:
\[
    (\hat U_b,\hat{\delta m}_{a\rightarrow b})
    =
    P_\theta(U_a^{\mathrm{skip}},m_a^{\mathrm{skip}},a,b,t),
\]
\[
    \tilde Z_b
    =
    \exp\!\left(\operatorname{clip}\left(
    m_a^{\mathrm{skip}}+s_m\hat{\delta m}_{a\rightarrow b}^{\mathrm{norm}},
    3,8\right)\right)\hat U_b .
\]
The backbone then resumes from layer $b+1$ and predicts
$\hat v_{\mathrm{skip}}$. The auxiliary task-aware loss is
\[
    \mathcal{L}_{\mathrm{skip-FM}}
    =
    \|\hat v_{\mathrm{skip}}-(x_0-x_1)\|_2^2.
\]
The recommended initial setting for SiT-B/2 with P-Small is
\[
\begin{array}{ll}
p_{\mathrm{skip}} & 0.10 \\
\lambda_{\mathrm{skip}} & 0.10 \\
d_{\mathrm{skip}} & \text{truncated-normal on }[1,10]\text{ with loc }3\text{ and sigma }2 \\
\text{warmup} & 5000\ \text{steps} \\
\text{source detach} & \text{enabled}
\end{array}
\]
The full objective in this mode is
\[
    \boxed{
    \mathcal{L}
    =
    \mathcal{L}_{\mathrm{FM}}
    +0.5\,\mathcal{L}_{\mathrm{dir}}
    +0.25\,\mathcal{L}_{\mathrm{boot}}
    +0.375\,\mathcal{L}_{\mathrm{mag}}
    +0.1875\,\mathcal{L}_{\mathrm{boot-mag}}
    +0.1\,\mathcal{L}_{\mathrm{skip-FM}}
    }.
\]
Log $\mathcal{L}_{\mathrm{skip-FM}}$, the effective skip probability, sampled
gap, and sampled source/target layers. If full-backbone FID degrades, reduce
$p_{\mathrm{skip}}$ or $\lambda_{\mathrm{skip}}$ before shifting probability
mass toward longer gaps.

\section{EMA Magnitude Calibration}
In the default \texttt{direction} mode, magnitude is not predicted by the
shortcut model. For skip inference ablation in that mode only, maintain
\[
    R_{\mathrm{EMA}}\in\mathbb{R}^{(L+1)\times5\times P}.
\]
The timestep bin is
\[
    \tau(q)=\left\lfloor \frac{q}{10}\right\rfloor,\qquad
    \tau(q)\in\{0,1,2,3,4\}.
\]
For a minibatch, define the per-sample token magnitude
\[
    r_l^{(n)}(i)=\|Z_l^{(n)}(i)\|_2,\qquad
    l\in\{0,\ldots,L\},\quad i\in\{1,\ldots,P\}.
\]
For each layer $l$, bin $\tau$, and token $i$, compute the batch estimate
\[
    \bar r_{l,\tau,i}
    =
    \frac{
        \sum_{n=1}^{B}\mathbf{1}[\tau(q_n)=\tau]\,
        r_l^{(n)}(i)
    }{
        \max\left(1,\sum_{n=1}^{B}\mathbf{1}[\tau(q_n)=\tau]\right)
    }.
\]
Across devices, the numerator and denominator are summed before applying the
EMA update. If a bin has no samples in the global batch, that entry is left
unchanged. Otherwise:
\[
    R_{\mathrm{EMA}}[l,\tau,i]
    \leftarrow
    0.99\,R_{\mathrm{EMA}}[l,\tau,i]
    +0.01\,\bar r_{l,\tau,i}.
\]
Thus $R_{\mathrm{EMA}}[l,\tau,i]$ estimates
\[
    \mathbb{E}\!\left[\|Z_l(i)\|_2\,\middle|\,\tau(q)=\tau\right].
\]

At skip inference from layer $a$ to $b$, the predictor first produces only a
direction:
\[
    U_a=\operatorname{dir}(Z_a),\qquad
    Y_{a\rightarrow b}=P_\theta(U_a,a,b,t_q),\qquad
    \hat U_{a\rightarrow b}=\operatorname{dir}(Y_{a\rightarrow b}).
\]
The restored hidden state is then
\[
    \boxed{
    \tilde Z_b^{(n)}(i)=
    R_{\mathrm{EMA}}[b,\tau(q_n),i]\,
    \hat U_{a\rightarrow b}^{(n)}(i)
    }.
\]
This is the state passed back to the DiT before resuming from block $b+1$.
Without this calibration, skip sampling would feed unit-norm hidden directions
into later DiT blocks, which expect layer- and timestep-dependent hidden-state
magnitudes.

In \texttt{direction-magnitude} mode, this EMA calibration is not used for
skip sampling. Instead, the predictor restores each skipped state from its
own predicted log magnitude:
\[
    \hat Z_b^{(n)}(i)
    =
    \exp\!\left(
    \operatorname{clip}\left(m_a^{(n)}(i)+
    \hat{\delta m}_{a\rightarrow b}^{(n)}(i),3,8\right)
    \right)
    \hat U_{a\rightarrow b}^{(n)}(i).
\]

\section{Optimization and Metrics}
The DiT uses AdamW with backbone learning rate from the main config and
weight decay $0.01$, excluding biases, normalization parameters, and
embedding tables. The predictor uses AdamW with learning rate $10^{-4}$,
weight decay $0.1$, betas $(0.9,0.999)$, and gradient clipping at $1.0$.

Log the training losses and cosine similarities:
\[
    \cos_{\mathrm{dir}}=
    \frac{1}{BP}\sum_{n,i}
    \hat U_{a\rightarrow b}^{(n)}(i)^\top U_b^{(n)}(i),
\]
\[
    \cos_{\mathrm{boot}}=
    \frac{1}{BP}\sum_{n,i}
    \hat U_{a\rightarrow b\rightarrow c}^{(n)}(i)^\top
    \hat U_{a\rightarrow c}^{\mathrm{ema},(n)}(i).
\]
For inference ablations, also report restored-state relative error, output
deviation, FID, wall-clock speedup, and the FID--speed curve.
In \texttt{direction-magnitude} mode, also log direct-pair magnitude errors:
\[
    \operatorname{ClipRate}_{\delta m}
    =
    \mathbb{P}_{n,i}\left(|m_b^{(n)}(i)-m_a^{(n)}(i)|>s_m\right).
\]
If this rate is above roughly $1\%$ for SiT-B/2, increase $s_m$ from $3$ to
$3.5$.
\[
    \operatorname{MAE}_{\delta m}=
    \mathbb{E}_{n,i}\left|\hat{\delta m}_{a\rightarrow b}^{(n)}(i)
    -\delta m_{a\rightarrow b}^{\mathrm{tgt},(n)}(i)\right|,
\]
\[
    \operatorname{RMSE}_{\delta m}
    =
    \sqrt{
    \mathbb{E}_{n,i}
    \left(\hat{\delta m}_{a\rightarrow b}^{(n)}(i)
    -\delta m_{a\rightarrow b}^{\mathrm{tgt},(n)}(i)\right)^2
    },
\]
\[
    \operatorname{RatioError}
    =
    \mathbb{E}_{n,i}
    \exp\!\left(
    \left|\hat{\delta m}_{a\rightarrow b}^{(n)}(i)
    -\delta m_{a\rightarrow b}^{\mathrm{tgt},(n)}(i)\right|
    \right).
\]
For diagnosis, \texttt{--shortcut-debug-gap-logs} evaluates predictor losses on
fixed representative pairs every train step, independent of the randomly
sampled training pair. The logged pairs are
\[
    1\!\rightarrow\!3,\quad 3\!\rightarrow\!5,\quad 5\!\rightarrow\!7,\quad
    10\!\rightarrow\!12,\quad 1\!\rightarrow\!5,\quad 3\!\rightarrow\!7,\quad
    6\!\rightarrow\!12,\quad 1\!\rightarrow\!10,\quad 3\!\rightarrow\!12.
\]
For each pair the debug metrics are direction loss, magnitude loss, direction
cosine, and log-magnitude MAE.
The same flag also logs the average of those four metrics over all valid
source--target pairs for each gap $d\in\{1,\ldots,10\}$, e.g.
\texttt{train/debug\_gap\_3/loss\_dir}.

\section{Skip-FID Evaluation}
In addition to full-backbone FID, evaluate a simpler skip sampler that makes
one shortcut jump only:
\[
    3\rightarrow7.
\]
At source layer $3$, the predictor estimates the target direction:
\[
    \hat U_{3\rightarrow7}
    =
    \operatorname{dir}(P_{\bar\theta}(\operatorname{dir}(Z_3),3,7,t_q)).
\]
In the default direction-only mode, restore the target-layer magnitude with
the L2 EMA:
\[
    \tilde Z_7(i)=
    R_{\mathrm{EMA}}[7,\tau(q),i]\,
    \hat U_{3\rightarrow7}(i).
\]
This is exactly the magnitude calibration above: the raw predictor output
$Y$ is not used as a full hidden state; only its direction is used, and the
target magnitude comes from $R_{\mathrm{EMA}}$.
In \texttt{direction-magnitude} mode, skip-FID instead uses predicted log
magnitude and no EMA calibration:
\[
    \tilde Z_7(i)=
    \exp\!\left(\operatorname{clip}\left(
    m_3(i)+\hat{\delta m}_{3\rightarrow7}(i),3,8
    \right)\right)\hat U_{3\rightarrow7}(i).
\]
The sampler then skips the true intervening blocks and resumes from the block
after the restored target state. The validation metrics are logged as
\texttt{val/FID\_skip\_3to7} and \texttt{val/sFID\_skip\_3to7}.

\section{Experimental Plan}
Run the method in four stages so implementation bugs, capacity limits, and
skip-quality tradeoffs are separated.

\paragraph{Stage 1: Pipeline debug.}
Train with P-Tiny for short runs. The goal is only to verify that the joint
loss is numerically stable, EMA teacher updates, $R_{\mathrm{EMA}}$ fills all
five timestep bins, and the raw predictor norm remains reasonable:
\[
    \mathbb{E}_{n,i}\|Y_{a\rightarrow b}^{(n)}(i)\|_2,\qquad
    \min_{n,i}\|Y_{a\rightarrow b}^{(n)}(i)\|_2,\qquad
    \max_{n,i}\|Y_{a\rightarrow b}^{(n)}(i)\|_2 .
\]
Since $\|U_a(i)\|_2=1$, these values should not collapse near zero or grow to
large magnitudes before direction normalization.

\paragraph{Stage 2: Main model.}
Train P-Base as the balanced setting for the primary result. Track
$\mathcal{L}_{\mathrm{FM}}$, $\mathcal{L}_{\mathrm{dir}}$,
$\mathcal{L}_{\mathrm{boot}}$, $\cos_{\mathrm{dir}}$, and
$\cos_{\mathrm{boot}}$. Evaluate both full-backbone FID and skip-FID at the
same checkpoint cadence.

\paragraph{Stage 3: Capacity sweep.}
If P-Base underfits long gaps, run P-Large with the same optimizer settings.
Compare long-gap direct cosine and skip-FID against P-Base for skips such as
$1\rightarrow3$, $3\rightarrow5$, and longer diagnostic pairs like
$0\rightarrow6$ or $0\rightarrow12$.

\paragraph{Stage 4: Quality--speed curve.}
For selected checkpoints, report full FID/sFID and skip FID/sFID together
with wall-clock sampling time. The useful operating point is the largest
speedup whose FID degradation remains acceptable.

\end{document}
